Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-19 latex/main.tex
Update methods/comparison with UShER

\subsubsection*{Ancestral Recombination Graphs}

%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
the term ``Ancestral Recombination Graph'' has come to be used
more generally to describe not just realizations of the model,
but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
While there is some subtlety in the details \cite{Wong2024},
we can think of an ARG as being any graph
that encodes the reticulate genetic genealogy of a sample of sequences
under the influence of recombination.

@@

\subsubsection*{ARG postprocessing}
The first step in ARG postprocessing is to examine putative recombinants to
determine whether a more parsimonious explanation is possible. This is done 
by rematching the inferred recombinant sequence against the ARG pruned back
to include only samples before the date of the recombinant.

@@

\subsubsection*{Comparison with UShER}
To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
properties but is more general (see the section ``Ancestral Recombination
Graphs``).
We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
from Figshare
and converted the ARG topology. We then downloaded the protobuf formatted
tree (tree.all\_viridian.202409.pb.gz), and exported the mutations to 
nucleotide format using \texttt{matUtils summary --translate}
(usher version 0.6.3), and translated
these mutations into tskit format. The resulting tszip file 
(usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
mutations and requires 29MB of storage space.
We validated that the data was correctly encoded by comparing the 
variant data for the UShER tskit tree sequence against our alignments.
There was an exact match between the non-ambiguous nucleotide calls 
at 27469 sites. At the remaining 39 sites there were a significant 
number of gap characters (deletion calls), suggesting that the mismatches
could be due to differences in the alignments used as input to UShER.
See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
for details.

We then computed intersection ARGs for sc2ts and UShER by 
``simplifying''\cite{Kelleher2018efficient,Wong2024} 
both to the 2,475,418 samples and 27,507 sites that are shared.

All conversions were performed as part of the reproducible 
postprocessing pipeline described in [section?]. The full UShER
tree and the intersection files are available for download on Zenodo
in tszip format.

@@

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an ``origination node" for a focal Pango lineage,
defined as the oldest internal node in the ARG
whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
and where the internal node is itself designated as the focal Pango.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e.\ the Pango X lineages) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

We compare topology in the simplified ARG against the UShER tree, built from the Viridian dataset.
To remove time traveling samples from the UShER tree,
we reduce it to 2475419 samples shared with the main sc2ts ARG
and date the internal nodes by running tsdate,
with no rescaling and the root node constrained to the established date of the UShER root (2020-01-16).
To compare against the ARG, we simplify this UShER tree down to the 1667 representative ARG samples.

>>>>

\subsubsection*{Ancestral Recombination Graphs}
\label{sec:args}
%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
the term ``Ancestral Recombination Graph'' has come to be used
more generally to describe not just realizations of the model,
but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
While there is some subtlety in the details \cite{Wong2024},
we can think of an ARG as being any graph
that encodes the reticulate genetic genealogy of a sample of sequences
under the influence of recombination.

@@

\subsubsection*{ARG postprocessing}
\label{sec:arg_postprocessing}
The first step in ARG postprocessing is to examine putative recombinants to
determine whether a more parsimonious explanation is possible. This is done 
by rematching the inferred recombinant sequence against the ARG pruned back
to include only samples before the date of the recombinant.

@@

\subsubsection*{Comparison with UShER}
To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
properties but is more general 
(see the \nameref{sec:args} section).
We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
and converted the tree topology to tskit node and edge table 
descriptions. We then downloaded the protobuf formatted
tree (tree.all\_viridian.202409.pb.gz), exported the mutations to 
nucleotide format using \texttt{matUtils summary --translate}
(usher version 0.6.3), and translated
these mutations into tskit site and mutation tables. 
The resulting tszip file 
(usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
mutations and requires 29MB of storage space.
We validated that the data was correctly encoded by comparing the 
variant data for the UShER tskit tree sequence against our alignments.
There was an exact match between the non-ambiguous nucleotide calls 
at 27,469 sites. 
% Usher tree has 27,508 vs sc2ts 29,893, or 2385 fewer sites 
At the remaining 39 sites for which the UShER tree 
had mutations there were a significant 
number of gap characters (deletion calls), suggesting that the mismatches
could be due to differences in the alignments used as input to UShER.
See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
for details.

We then computed intersection ARGs for sc2ts
(sc2ts\_viridian\_inter\_v1.2.trees.tsz) and UShER 
(usher\_viridian\_inter\_v1.2.trees.tsz) 
by 
simplifying\cite{Kelleher2018efficient,Wong2024} 
both to the 2,475,418 samples and 27,507 sites that are shared
(the UShER tree has mutations at 27,508 sites and sc2ts ARG has 29,893).
As there was no documented means of deriving node dates from 
the UShER tree, and we wish to visually compare 
the phylogenetic backbones, we
also dated the intersection tree by fixing
the sample dates using the ``Date\_tree'' metadata field, 
and using tsdate (version 0.2.3) to date internal nodes.
We did not date the main UShER tree in this way because
of the time-travelling samples present.

All conversions were performed as part of the reproducible 
pipeline described in the \nameref{sec:arg_postprocessing} section. 
The full UShER
tree and the intersection files are available for download 
in tszip format (see \nameref{sec:data_and_code_availability}).

@@

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an ``origination node'' for each Pango lineage,
defined as the oldest node in the ARG with that Pango designation
whose descendants in any tree include more than 50\% of the Pango samples of
the focal type.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e.\ the Pango X lineages) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

We compare topology of the simplified ARG against the intersection UShER tree
(which has been dated) simplified down to the same 1667 samples.
